---
title: "Capital Asset Pricing Model"
author: "Team Chen, Edwell, Praise"
format: 
  html:
    toc: true          
    toc-depth: 3       
    warning: false
    error: false
    message: false
editor: visual
---

## Stage 1: Load data

```{r}
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(readr))
```

```{r}
df <- read_csv("M:/UW QF/Sem 2/Reproducible Research/New folder/cleaned_data.csv", show_col_types = FALSE )

```

### Checking the first few rows

```{r}
head(df)
```

### Converting Date column to Date type

```{r}
df$Date <- as.Date(df$Date)
```

## Stage 2: Prep and normalizing returns

```{r}
# Select columns of interest: Date + selected stocks + market returns
returns_df <- df %>% select(Date, BA, T, MGM, AMZN, IBM, TSLA, GOOG, Ri, Rm)

# Calculate cumulative index (base 100) from daily returns
cumulative_df <- returns_df
cumulative_df[,-1] <- apply(returns_df[,-1], 2, function(x) cumprod(1 + x) * 100)

# Preview cumulative returns
head(cumulative_df)

```

## Stage 3: Interactive plot of cumulative stock performance

```{r}
interactive_plot <- function(df, title) {
  fig <- plot_ly()
  
  for (col_name in colnames(df)[-1]) {
    fig <- fig %>%
      add_trace(x = df$Date,
                y = df[[col_name]],
                type = 'scatter',
                mode = 'lines',
                name = col_name)
  }
  
  fig <- fig %>%
    layout(title = title,
           xaxis = list(title = "Date"),
           yaxis = list(title = "Index Value (Base = 100)"))
  
  return(fig)
}

# Generate the interactive plot
interactive_plot(cumulative_df, "Stock Performance Indexed to 100")

```

## Stage 4: Average daily returns

```{r}
avg_daily_returns <- df %>%
  select(-Date) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE)))

print(avg_daily_returns)

```

## Stage 5: Plot - Single Stock (AAPL) vs Market Returns

```{r}
ggplot(df, aes(x = Rm, y = Ri)) +
  geom_point(color = "darkgreen") +
  labs(
    title = "AAPL vs Market Returns",
    x = "S&P 500",
    y = "AAPl"
  ) +
  theme_minimal()

```

## Stage 6: Linear regression and regression plot

```{r}
# Convert returns to percentage
df$Ri_pct <- df$Ri * 100
df$Rm_pct <- df$Rm * 100

# Fit linear regression model
model <- lm(Ri_pct ~ Rm_pct, data = df)

# Extract coefficients
beta <- coef(model)[["Rm_pct"]]
alpha <- coef(model)[["(Intercept)"]]

cat("Beta for AAPL stock is =", round(beta, 3), "and alpha is =", round(alpha, 3), "\n")

# Plot scatter with regression line
ggplot(df, aes(x = Rm_pct, y = Ri_pct)) +
  geom_point(color = "blue") +
  geom_abline(intercept = alpha, slope = beta, linetype = "dashed", color = "red") +
  labs(
    title = "Scatter Plot: AAPL vs S&P500 Returns",
    x = "Market Return (S&P 500, %)",
    y = "AAPL Return (%)"
  ) +
  theme_minimal()

```

## Stage 7: Annualized market return and CAPM expected returns

```{r}
# Define stock columns to analyze (update this list as per your data)
stock_columns <- c("BA", "T", "MGM", "AMZN", "IBM", "TSLA", "GOOG", "Ri") 

# Ensure market returns in percentage
df$Rm_pct <- df$Rm * 100

# Risk-free rate (annual, in %)
rf <- 0.66

# Annualize market return (mean daily return * 252 trading days * 100 to convert to %)
annual_rm <- mean(df$Rm, na.rm = TRUE) * 252 * 100
cat("Annualized market return (S&P500):", round(annual_rm, 3), "%\n")

# Data frame to store all results
results <- data.frame(Stock=character(), Beta=numeric(), Alpha=numeric(), ExpectedReturn=numeric(), stringsAsFactors=FALSE)

for (stock in stock_columns) {
  
  # Create daily returns in percent for the stock
  stock_pct_col <- paste0(stock, "_pct")
  df[[stock_pct_col]] <- df[[stock]] * 100
  
  # Fit linear regression: Stock returns ~ Market returns
  model <- lm(df[[stock_pct_col]] ~ df$Rm_pct)
  beta <- coef(model)[2]
  alpha <- coef(model)[1]
  
  # Calculate CAPM expected return
  expected_return <- rf + beta * (annual_rm - rf)
  
  # Save to results
  results <- rbind(results, data.frame(Stock = stock, Beta = beta, Alpha = alpha, ExpectedReturn = expected_return))
  
  # Plot scatter + regression line for each stock
  plot_data <- data.frame(Market = df$Rm_pct, Stock = df[[stock_pct_col]])
  p <- ggplot(plot_data, aes(x = Market, y = Stock)) +
    geom_point(color = "blue", alpha=0.5) +
    geom_abline(intercept = alpha, slope = beta, color = "red") +
    labs(title = paste("Scatter Plot & Regression:", stock, "vs S&P500"),
         x = "Market Return (%)",
         y = paste(stock, "Return (%)")) +
    theme_minimal()
  
  print(p)
  
  # Print beta, alpha, expected return
  cat(sprintf("Stock: %s | Beta: %.3f | Alpha: %.3f | Expected CAPM Return: %.3f%%\n", stock, beta, alpha, expected_return))
}

# Round results for better readability
results$Beta <- round(results$Beta, 3)
results$Alpha <- round(results$Alpha, 3)
results$ExpectedReturn <- round(results$ExpectedReturn, 3)

print(results)


```

## Stage 8: Portfolio expected returns based on CAPM

```{r}
# Calculate expected returns for each stock using CAPM formula
rf <- 0.66
rm <- round(mean(df$Rm, na.rm = TRUE) * 252 * 100, 1)

ER <- setNames(numeric(length(stock_columns)), stock_columns)

for (stock in stock_columns) {
  beta <- results$Beta[results$Stock == stock]
  ER[stock] <- rf + beta * (rm - rf)
  cat(sprintf("Expected Return Based on CAPM for %s is %.3f%%\n", stock, ER[stock]))
}

# Example: Equal weights portfolio (assuming 8 stocks)
portfolio_weights <- rep(1/length(stock_columns), length(stock_columns))

# Portfolio expected return
ER_portfolio <- sum(ER * portfolio_weights)

cat(sprintf("Expected Return for equal-weighted portfolio is %.3f%%\n", ER_portfolio))

```

## Stage 9: Validation

```{r}
# Stage 9: Validation statistics for all stocks CAPM models

library(broom)  # for tidy()

# Prepare data in long format for modeling
returns_long <- df %>%
  select(Date, all_of(stock_columns), Rm_pct) %>%
  pivot_longer(cols = all_of(stock_columns), names_to = "Stock", values_to = "Return") %>%
  # Ensure both Return and Rm_pct have no NA/Inf
  filter(!is.na(Return), !is.na(Rm_pct), !is.infinite(Return), !is.infinite(Rm_pct))

# Run linear regression for each stock
models <- returns_long %>%
  group_by(Stock) %>%
  summarise(
    model = list(lm(Return ~ Rm_pct, data = cur_data_all()))
  )

# Extract model statistics in a tidy dataframe
validation_stats <- models %>%
  mutate(
    tidy = map(model, broom::tidy),
    glance = map(model, broom::glance)
  ) %>%
  unnest(tidy) %>%
  filter(term == "Rm_pct") %>%
  select(Stock, estimate, std.error, statistic, p.value) %>%
  rename(
    Beta = estimate,
    StdError = std.error,
    Tvalue = statistic,
    Pvalue = p.value
  ) %>%
  left_join(
    models %>%
      mutate(glance = map(model, broom::glance)) %>%
      unnest(glance) %>%
      select(Stock, r.squared, adj.r.squared),
    by = "Stock"
  ) %>%
  arrange(Stock)

# Print the validation statistics summary
print(validation_stats)

# Optional: save to CSV
# write_csv(validation_stats, "capm_validation_stats.csv")

```

All betas are positive and statistically significant, meaning all these stocks have returns positively related to market returns, consistent with CAPM assumptions.

Magnitude of betas are small (\~0.007 to 0.016) — this may be because returns are expressed in daily % terms and/or your market returns scale is low. The relationship exists but daily beta is small.

R-squared values vary from about 0.15 to 0.5, indicating the market explains between 15% to 50% of return variance. Stocks like IBM, GOOG have stronger market relationships, while TSLA has the weakest.

The very low p-values (\< 0.0001) confirm the relationships are highly statistically significant for all stocks.
